### 2021-05-06 doc/thesis/chapters/dsp.tex
update

@@

Frequency modulation (FM) is a widely used standard to transmit data streams.
The probably best known usecase therefore is commercial broadcast radio, where an audio stream is transmitted.
Devices to receive these streams are available for low prices to the public.

@@

Formula \eqref{equ:fm_func} describes a frequency modulated signal with a generic message signal $m(t)$.
A widely used example application of FM is broadcast radio, where audio streams are transmitted.
An audio stream can be described as a cosine wave.
Therefor, the information signal that is to be transmitted in broadcast radio can be represented as a cosine wave in the form of
\begin{equation}
  m(t) = A_m\ cos(2 \pi f_m t).
\end{equation}
By inserting this message signal into the generic FM formula \eqref{equ:fm_func}, the final equation for an FM signal transforms into the following form.
\begin{equation}
  y(t) = A_c\ cos \Big(2 \pi f_c t + \beta\ sin(2 \pi f_m t)\Big )
\end{equation}
Several formulas, as well as their derivations are described according to \cite{ref_FM_Maths_Info_1}\cite{ref_FMMaths2}.

@@

Each channel may allocate a bandwidth of 200 kHz.
Within one of these channels, the maximum deviation from the center frequency shall not exceed 75 kHz.
This leaves a guard band of 25 kHz on either side, to minimize interference with adjacent channels.
Center frequencies may be allocated on multiples of 100 kHz.
However, neighboring channels need to be considered therefor, since this may result in an overlap in the spectrum.\\

Fig.\ref{fig_channel_baseband_freqs} shows the frequency spectrum of a single channel, located at its carrier- or center frequency.
For a better overview, only the positive spectum part is shown.
The negative part is an exact mirror, because the transmitted and received signals from the antenna are real signals - they do not have an imaginary part.
The spectrum consists of multiple parts, which are described in the following paragraphs.

@@

\subsubsection{Mono Audio Part}

@@

\subsubsection{Pilot Tone}

@@

\subsubsection{Stereo Audio Difference}

@@

\subsubsection{Additional Services}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algorithms for Digital FM Demodulation}
  %see literature/FmDemodulator.pdf (Sect. 3.3)
  %see literature/00476180 Digital FM Demodulator for FM, TV, and Wireless.pdf (Sect. II and III)

@@

Different methods may be implemented for the envelope detector.
For example, an `Asynchronous Full-Wave Envelope Detector' can be used.
Therefore, the previous method only needs to swap the thresholding unit with an unit that calculates the absolut value.
In that architecture, a full-wave rectification is performed on the signal, which significantly improves the envelope detection accuracy \cite{ref_envelope_detector}.
The envelope is followed more exactly, because of the higher power that is available in the signal, since the entire signal is taken into account, and not only the positive half, as in the previous half-wave method.

@@

This chapter describes the demodulation of the information content of an FM broadcast channel, as it is used in commercial FM radio broadcasting.
The assumption here is, that the actual FM demodulation, as described in chapter \ref{sec_fm_sig_demod} is already done correctly.
A frequency spectrum of an FM channel is shown in Fig.\ref{fig_channel_baseband_freqs} above.

explain multiplex signal somewhere

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mono}
\label{subsec:demod_mono}
  The first part of the channel frequency spectrum is the mono audio signal. It is summation signal of the left and right audio channel.
  A mono receiver can thus simply replay the so-called multiplex signal, which is generated by the FM demodulation described in chapter \ref{sec_fm_sig_demod}. The 19 kHz pilot tone will not be audible by most people, because it is outside of the range of a human ear. Besides that, any used speaker may also be unable to produce a frequency in this range. This explanation applies to several higher frequency components in the spectrum as well.
  In case the multiplex signal should be stored it in a file with 44.1 kHz samplerate, a lowpass with a cutoff frequency of 15 kHz and sufficient attenutation before 19 kHz needs to be inserted before the decimation and file output.

\subsection{Stereo}
  Demodulating an FM channel to recover a stereo audio signal requires a more sophisticated approach.
  In the channel frequency spectrum, there are two main parts of the audio signal - the sum and difference of the left and right audio channel signals.
  To combine these to a stereo signal, the following equations can be applied.

  \begin{equation}
    \begin{split}
      L = (L+R) + (L-R) = (2)L \\
      R = (L+R) - (L-R) = (2)R
      \label{equ_stereo_from_sum_diff}
    \end{split}
  \end{equation}

  The block diagram in Fig.\ref{fig_bd_stereo_demod} illustrates the signal processing chain for stereo audio. In the block diagram, $x(t)$ represents the multiplex signal from the FM demodulator. Signals $x_l(t)$ and $x_r(t)$ describe the left and right audio signal, respectively.

@@

  The central branch selects the pilot tone at 19 kHz with a bandpass filter.
  This bandpass needs to have a frequency response that is sharp enough to have a sufficient attenuation at $\pm$4 kHz, since this is where the summation and difference signals frequencies are allocated.
  Afterwards, the pilot tone frequency is doubled via a PLL to achieve a 38 kHz to later demodulate the difference signal phase-coherently.
  The PLL lock indicator can be re-used as an indicator, whether a pilot tone is existent.
  In the upper branch, a bandpass filter selects the difference signal with ranges from 23 to 53 kHz.
  It is then modulated to baseband by the previously generated 38 kHz subcarrier.
  Subsequently, a lowpass limits the signal to the audio signal bandwidth of 15 kHz.
  The lower branch lowpass-filters the summation signal.
  The rightmost part in the diagram performs the combination of summation and difference signals, according to Eq.\eqref{equ_stereo_from_sum_diff}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Oversampling / Sample Rate Reduction}
%TODO: explain DSP functions that are actually used in Matlab FM demod
Example: \\
In case the multiplex signal needs to be stored in a file with 44.1 kHz samplerate, in order to avoid aliasing, a lowpass with a cutoff frequency of 15 kHz and sufficient attenutation before 19 kHz needs to be inserted before the decimation and file output.

@@

Frequency modulation (FM) is a widely used standard to transmit data streams.
The probably best known usecase therefor is commercial broadcast radio, where an audio stream is transmitted.
Devices to receive these streams are available for low prices to the public.

@@

Formula \eqref{equ:fm_func} describes a frequency modulated signal with a generic message signal $m(t)$.
A widely used example application of FM is broadcast radio, where audio streams are transmitted.
An audio stream can be described as a cosine wave.
Therefore, the information signal that is to be transmitted in broadcast radio can be represented as a cosine wave in the form of
\begin{equation}
  m(t) = A_m\ cos(2 \pi f_m t).
\end{equation}
By inserting this message signal into the generic FM formula \eqref{equ:fm_func}, the final equation for an FM signal transforms into the following form.
\begin{equation}
  y(t) = A_c\ cos \Big(2 \pi f_c t + \beta\ sin(2 \pi f_m t)\Big )
\end{equation}
Several formulas, as well as their derivations are described according to \cite{ref_FM_Maths_Info_1}\cite{ref_FMMaths2}.

@@

Each channel may allocate a bandwidth of 200 kHz.
Within one of these channels, the maximum deviation from the center frequency shall not exceed 75 kHz.
This leaves a guard band of 25 kHz on either side, to minimize interference with adjacent channels.
Center frequencies may be allocated on multiples of 100 kHz.
However, neighboring channels need to be considered therefore, since this may result in an overlap in the spectrum.\\

Fig.\ref{fig_channel_baseband_freqs} shows the frequency spectrum of a single channel, located at its carrier- or center frequency.
For a better overview, only the positive spectum part is shown.
The negative part is an exact mirror, because the transmitted and received signals from the antenna are real signals - they do not have an imaginary part.
The spectrum consists of multiple parts. Therefore, it is also referred to as `multiplex signal' in the literature. Several parts are described in the following paragraphs.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Mono Audio Part}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Pilot Tone}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Stereo Audio Difference}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Additional Services}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algorithms for Digital FM Demodulation}
\label{sec:algorithms-for-digital-fm-demodulation}
  %see literature/FmDemodulator.pdf (Sect. 3.3)
  %see literature/00476180 Digital FM Demodulator for FM, TV, and Wireless.pdf (Sect. II and III)

@@

Different methods may be implemented for the envelope detector.
For example, an `Asynchronous Full-Wave Envelope Detector' can be used.
Therefor, the previous method only needs to swap the thresholding unit with an unit that calculates the absolut value.
In that architecture, a full-wave rectification is performed on the signal, which significantly improves the envelope detection accuracy \cite{ref_envelope_detector}.
The envelope is followed more exactly, because of the higher power that is available in the signal, since the entire signal is taken into account, and not only the positive half, as in the previous half-wave method.

@@

This chapter describes the demodulation of the information content of an FM broadcast channel, as it is used in commercial FM radio broadcasting.
The assumption here is, that the actual FM demodulation, as described in chapter \ref{sec:algorithms-for-digital-fm-demodulation} is already done correctly.
A frequency spectrum of the multiplex signal of an FM channel is shown in Fig.\ref{fig_channel_baseband_freqs}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mono Audio}
\label{subsec:demod_mono}

The first part of the channel frequency spectrum is the mono audio signal.
It is the summation signal of the left and right audio channel.
Thus, it contains enough information to replay an audio stream in mono quality.
A mono receiver can thus simply replay the so-called multiplex signal that is produced by the FM demodulator, without any further filtering or processing.
The 19 kHz pilot tone, and any higher frequency components, will not be audible by most people, because it is outside of the range of a human ear.
Besides that, any used speaker may also be unable to produce a frequency in this range.\\

Consequently, in order to demodulate this spectral part, a lowpass filter is required.
The filter needs to have a cutoff frequency of 15 kHz, to allow the entire mono audio spectrum to pass.
Sufficient attenutation must be available before 19 kHz, in order to suppress the pilot tone.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Stereo Audio}

The demodulation of an FM channel to recover a stereo audio signal, requires a more sophisticated approach and multiple steps in signal processing.
Two parts need to be combined in order to achieve stereo.
The sum and the difference of the left and right audio channel signals - the mono and the stereo difference part, respectively.
The following equations need to be applied, to compute a left and a right channel.

\begin{equation}
  \begin{split}
    (L+R) + (L-R) = (2)L \\
    (L+R) - (L-R) = (2)R
    \label{equ_stereo_from_sum_diff}
  \end{split}
\end{equation}

where $(L+R)$ represents the mono part and $(L-R)$ the stereo difference.\\

The block diagram in Fig.\ref{fig_bd_stereo_demod} illustrates the signal processing chain for stereo audio.
In the block diagram, $x(t)$ represents the multiplex signal from the FM demodulator.
The signals $x_l(t)$ and $x_r(t)$ describe the left and right audio signal, respectively.\\

@@

The central branch selects the pilot tone at 19 kHz with a bandpass filter.
This bandpass needs to have a frequency response that is sharp enough to have a sufficient attenuation at $\pm$4 kHz, since this is where the mono and difference signals' frequency spectra are allocated.
Afterwards, the pilot tone frequency is doubled to achieve a 38 kHz subcarrier.
This guarantees a phase-coherency between pilot and subcarrier, which is required to correctly shift the difference signal down to baseband.

The frequency doubling can be implemented with multiple methods.
The pilot tone can be modulated by simply multiplying it with itself.
Another option is to use a PLL and configure it to generate the double frequency at its output.
The PLL lock indicator can then be re-used as an indicator, whether a pilot tone is existent.\\

In the upper branch, a bandpass filter selects the difference signal with ranges from 23 to 53 kHz.
It is then modulated to baseband by the previously generated 38 kHz subcarrier.
Subsequently, a lowpass limits signal bandwidth to 15 kHz, which removes the generated modulation artifacts.\\

The lower branch lowpass-filters the summation, or mono signal, as described in paragraph \ref{subsec:demod_mono}.\\

The rightmost part in the diagram performs the combination of the mono and the difference signal, according to Eq.\eqref{equ_stereo_from_sum_diff}.
